Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States

### 2024-04-18 paper/paper.qmd
Finished model and results section.
Completed the model section and include information regarding the results that can be derived as a result.

# Package loading
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)

@@

cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
```

@@

```{r}
#| label: tbl-incarceration-data
#| tbl-cap: Top 10 countries by incarceration rate
#| echo: false

raw_incarceration_data |>
  arrange(desc(incarcerationRatesByCountry_ratePer100k)) |>
  head(10) |>
  select(country, incarcerationRatesByCountry_ratePer100k, incarcerationRatesByCountry_totIncarcerated) |>
  kable(
    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated")
  )
```

@@

```{r}
#| label: tbl-poverty-data
#| tbl-cap: Top 5 countries by poverty rate
#| echo: false

raw_poverty_data |>
  arrange(desc(PovertyRateValue)) |>
  head(5) |>
  select(country, PovertyRateValue, PovertyRateValueDataYear) |>
  kable(
    col.names = c("Country", "Poverty Rate", "Year Data was Collected")
  )
```

@@

```{r}
#| label: tbl-violent-crime-data
#| tbl-cap: Sample of 5 countries and their respective rates of various violent crimes (per 100k population)
#| echo: false

raw_violent_crime_data |>
  head(5) |>
  select(country, violentCrimeRatesByCountry_homRate, violentCrimeRatesByCountry_assault, violentCrimeRatesByCountry_sexViolence, violentCrimeRatesByCountry_kidnap, violentCrimeRatesByCountry_robbery) |>
  kable(
    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate")
  )
```

@@

```{r}
#| label: tbl-unemployment-data
#| tbl-cap: Top 5 countries by highest unemployment rate (as per the World Bank)
#| echo: false

raw_unemployment_data |>
  arrange(desc(UnemploymentRateWorldBank)) |>
  head(5) |>
  select(country, UnemploymentRateWorldBank,UnemploymentRateWorldBankDataYear) |>
  kable(
    col.names = c("Country", "Unemployment Rate", "Year Data was Collected")
  )
```

@@

```{r}
#| label: tbl-education-data
#| tbl-cap: Top 5 countries by education (World Top 20 (2024))
#| echo: false

raw_education_data |>
  arrange(educationRankingsByCountry_WT20rank2024) |>
  head(5) |>
  select(country, educationRankingsByCountry_WT20rank2024) |>
  kable(
    col.names = c("Country", "Education Rank")
  )
```

@@

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the sum. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, it was instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective. @tbl-cleaned-data

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Top 5 countries by incarceration rate in the finalized cleaned dataset
#| echo: false

cleaned_data |>
  arrange(desc(incarceration_rate)) |>
  head(5) |>
  kable(
        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate")
  )
```

@@

@fig-poverty-incarceration showcases all observations by poverty rate, highlighting the aforementioned countries as well. Similarly, @fig-unemployment-incarceration showcases all observations by unemployment rate and @fig-violent-crime-incarceration by violent crime rates. Lastly, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories. Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.

```{r}
#| label: fig-poverty-incarceration
#| fig-cap: Scatter plot of incarceration rates by poverty rates
#| echo: false
#| fig-pos: H

ggplot(cleaned_data, aes(x = poverty_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Poverty Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-unemployment-incarceration
#| fig-cap: Scatter plot of incarceration rates by unemployment rates
#| echo: false
#| fig-pos: H

ggplot(cleaned_data, aes(x = unemployment_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Unemployment Rate", y = "Incarceration Rate")
```

```{r}
#| label: fig-violent-crime-incarceration
#| fig-cap: Scatter plot of incarceration rates by violent crime rates
#| echo: false
#| fig-pos: H

@@

```{r}
#| label: tbl-summary-stats
#| tbl-cap: Mean and standard deviation for each variable
#| echo: false
#| fig-pos: H

@@

# Print the data frame as a table using kable
kable(data, col.names = c("Variable", "Mean", "SD"))
```

@@

Note, we employ [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.

@@

The above factors were each chosen due to the literature present regarding each of their impacts on incarceration. In particular, there is data to suggest that more impoverished individuals are more susceptible to being arrested, at least in the United States [@citePovertyImpact]. Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the US [@citeUnemploymentImpact]. There is also literature that suggests that education significantly reduces the likelihood of incarceration, with claims that incarceration rates monotonically decline with education [@citeEducationImpact]. As for the impact of violent crime rates on incarceration rates, most research generally tends to focus on the reverse, focusing on incarceration rates' impact on violent crime. As such, this paper seeks to explore the converse in order to gain a more well-rounded understanding of how both factors influence one another on the whole.

@@

@@

>>>>

# Package loading
library(tidyverse)
library(here)
library(ggplot2)
library(knitr)
library(modelsummary)
library(rstanarm)
library(kableExtra)

@@

cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))

# Read model
analysis_model <- readRDS(file = here("models/incarceration_model.rds"))

@@

```{r}
#| label: tbl-incarceration-data
#| tbl-cap: Top 10 countries by incarceration rate
#| echo: false
#| tbl-pos: H

raw_incarceration_data |>
  arrange(desc(incarcerationRatesByCountry_ratePer100k)) |>
  head(10) |>
  select(country, incarcerationRatesByCountry_ratePer100k, incarcerationRatesByCountry_totIncarcerated) |>
  kable(
    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated"), 
    booktabs = TRUE
  )
```

@@

```{r}
#| label: tbl-poverty-data
#| tbl-cap: Top 5 countries by poverty rate
#| echo: false
#| tbl-pos: H

raw_poverty_data |>
  arrange(desc(PovertyRateValue)) |>
  head(5) |>
  select(country, PovertyRateValue, PovertyRateValueDataYear) |>
  kable(
    col.names = c("Country", "Poverty Rate", "Year Data was Collected"), 
    booktabs = TRUE
  )
```

@@

```{r}
#| label: tbl-violent-crime-data
#| tbl-cap: Sample of 5 countries and their respective rates of various violent crimes (per 100k population)
#| echo: false
#| tbl-pos: H

raw_violent_crime_data |>
  head(5) |>
  select(country, violentCrimeRatesByCountry_homRate, violentCrimeRatesByCountry_assault, violentCrimeRatesByCountry_sexViolence, violentCrimeRatesByCountry_kidnap, violentCrimeRatesByCountry_robbery) |>
  kable(
    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate"), booktabs = TRUE
  ) |>
  column_spec(1, width = "1.8cm") |>
  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(6, width = "2.2cm")     # Adjust the width of the second column
```

@@

```{r}
#| label: tbl-unemployment-data
#| tbl-cap: Top 5 countries by highest unemployment rate (as per the World Bank)
#| echo: false
#| tbl-pos: H

raw_unemployment_data |>
  arrange(desc(UnemploymentRateWorldBank)) |>
  head(5) |>
  select(country, UnemploymentRateWorldBank,UnemploymentRateWorldBankDataYear) |>
  kable(
    col.names = c("Country", "Unemployment Rate", "Year Data was Collected"), 
    booktabs = TRUE
  )
```

@@

```{r}
#| label: tbl-education-data
#| tbl-cap: Top 5 countries by education (World Top 20 (2024))
#| echo: false
#| tbl-pos: H

raw_education_data |>
  arrange(educationRankingsByCountry_WT20rank2024) |>
  head(5) |>
  select(country, educationRankingsByCountry_WT20rank2024) |>
  kable(
    col.names = c("Country", "Education Rank"), booktabs = TRUE
  )
```

@@

In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation. It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the summation. As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking. This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective. @tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Top 5 countries by incarceration rate in the finalized cleaned dataset
#| echo: false
#| tbl-pos: H

cleaned_data |>
  arrange(desc(incarceration_rate)) |>
  head(5) |>
  kable(
        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate"), booktabs = TRUE
  ) |>
  column_spec(1, width = "2.2cm") |>
  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
  column_spec(6, width = "2.2cm")
```

@@

@fig-variable-incarceration showcases all observations by poverty rate, unemployment rate, and violent crime rates, highlighting the aforementioned countries as well. On the other hand, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories. Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.

```{r}
#| label: fig-variable-incarceration
#| fig-cap: Scatter plot of incarceration rates against each of the independent variables of interest.
#| fig-subcap: ["Poverty Rate", "Unemployment Rate", "Violent Crime Rate"]
#| echo: false
#| layout-ncol: 2
#| fig-pos: H

ggplot(cleaned_data, aes(x = poverty_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Poverty Rate", y = "Incarceration Rate")

ggplot(cleaned_data, aes(x = unemployment_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +  # Highlight specific countries
  theme_classic() +
  labs(x = "Unemployment Rate", y = "Incarceration Rate")

@@

```{r}
#| label: tbl-summary-stats
#| tbl-cap: Mean and standard deviation for each variable
#| echo: false
#| tbl-pos: H

@@

# Print the data frame as a table using kable
kable(data, col.names = c("Variable", "Mean", "SD"), booktabs = TRUE)
```

@@

Note, we employ `rstanarm` [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.

@@

The above factors were each chosen due to the literature present regarding each of their impacts on incarceration. In particular, there is data to suggest that more impoverished individuals are more susceptible to being arrested, at least in the United States [@citePovertyImpact]. Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the United States [@citeUnemploymentImpact]. There is also literature that suggests that education significantly reduces the likelihood of incarceration, with claims that incarceration rates monotonically decline with education [@citeEducationImpact]. As for the impact of violent crime rates on incarceration rates, most research generally tends to focus on the reverse, focusing on incarceration rates' impact on violent crime. As such, this paper seeks to explore the converse in order to gain a more well-rounded understanding of how both factors influence one another on the whole.

@@

## Model Coefficients

After running the regression based on the above model, we receive the following coefficient values as showcased in @tbl-incarcerationsummary.

```{r}
#| label: tbl-incarcerationsummary
#| tbl-cap: Model summary of the predicted impact of crime, education, unemployment, and poverty on incarceration rates.
#| echo: false
#| warning: false
#| message: false
#| tbl-pos: H

modelsummary(analysis_model)

We can see that, somewhat unexpectedly, increases in the poverty rate negatively correlate with incarceration rates, as do increases in unemployment, assuming all else held constant. On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant. As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the lowest category of education, "176-198", would see an increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing more consistent decreases near the highest ranked nations.

## Actual vs. Predicted Incarceration Rates

In order to get a better grasp of the predictive power of our model, we compare every nation's predicted incarceration rate, given their respective qualities, to their actual incarceration rate as showcased in @fig-actualvspredicted below.

```{r}
#| label: fig-actualvspredicted
#| tbl-cap: Comparison of actual and predicted incarceration rates.
#| echo: false
#| fig-pos: H

# Extract coefficient estimates
coefficients <- coef(analysis_model)

# Extract predictor variables (poverty, crime, unemployment, and education)
predictors <- cleaned_data |>
  select(poverty_rate, violent_crime_rate, unemployment_rate, education_category)

# Calculate the predicted incarceration rate for each nation
predicted_incarceration <- (
  coefficients["(Intercept)"] +                      
  coefficients["poverty_rate"] * predictors$poverty_rate +
  coefficients["violent_crime_rate"] * predictors$violent_crime_rate +
  coefficients["unemployment_rate"] * predictors$unemployment_rate +
  coefficients["education_category1-19"] * as.numeric(predictors$education_category == "1-19") +
  coefficients["education_category20-34"] * as.numeric(predictors$education_category == "20-34") +
  coefficients["education_category35-52"] * as.numeric(predictors$education_category == "35-52") +
  coefficients["education_category53-71"] * as.numeric(predictors$education_category == "53-71") +
  coefficients["education_category72-92"] * as.numeric(predictors$education_category == "72-92") +
  coefficients["education_category93-110"] * as.numeric(predictors$education_category == "93-110") +
  coefficients["education_category111-125"] * as.numeric(predictors$education_category == "111-125") +
  coefficients["education_category126-147"] * as.numeric(predictors$education_category == "126-147") +
  coefficients["education_category148-175"] * as.numeric(predictors$education_category == "148-175")
)

# Add predicted incarceration rate as a new column in cleaned_data
cleaned_data <- cleaned_data |>
  mutate(predicted_incarceration_rate = predicted_incarceration)

# Graph the resultant scatter plot
ggplot(cleaned_data, aes(x = predicted_incarceration_rate, y = incarceration_rate)) +
  geom_point(color = "darkgrey") +  # Add points for actual incarceration rates
  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +  # Add y=x line
  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
             aes(color = country), size = 1.5) +
  labs(x = "Predicted Incarceration Rate", y = "Actual Incarceration Rate") + 
  theme_classic()

In the above figure, each point represents a given nation's actual incarceration rate (on the y-axis), against their predicted incarceration rate (on the x-axis) based on our employed model. We include a dotted $y=x$ line for reference to showcase all the theoretical points where the actual incarceration rate of a nation equals the predicted incarceration rate. Note that any points below the dotted $y=x$ line signify nations whose predicted incarceration rates exceeds their actual incarceration rates, whereas points above the line signify nations whose actual exceeds their predicted.

Examining the results on the whole, it appears that the model is able to somewhat accurately predict a nation's incarceration rate, whenever that nation's actual incarceration rate is between 100 and 300. As a nation's actual incarceration rises above 300, the model loses its predictive power as it appears to significantly underestimate the nation's incarceration rate. On the other hand, with nations whose actual incarceration rates are below 100, it appears that model generally tends to overestimate the nation's incarceration rate.

We highlight the same 5 nations as done in @sec-data as they span the entire range of incarceration rates. We can see that, as observed, Iceland and Guinea-Bissau, who have the 5th lowest and lowest incarceration rates of 31 and 36, respectively, feature predicted incarceration rates that exceed their actual. El Salvador and the United States, who have the highest and 3rd highest incarceration rates of 1086 and 531, respectively, featured predicted incarceration rates far below their actual. On the other hand, Mexico, with an actual incarceration rate of 174, had a predicted incarceration rate almost exactly equal to its actual rate. Thus, these 5 nations effectively showcase where the model's predictive power is strongest - near the middle - and where it is weakest - on either extreme.

We discuss the implications of these results in @sec-discussion, and in particular, how this helps in understanding the United States' issue of mass incarceration.

@@

```{r}
#| label: fig-ppcheck-postprior
#| echo: false
#| fig-pos: H
#| message: false

pp_check(analysis_model) + theme_classic()

posterior_vs_prior(analysis_model) + theme_classic() + coord_flip()
```

```{r}
#| label: fig-rhat
#| echo: false
#| layout-ncol: 2
#| fig-pos: H

plot(analysis_model, "rhat")
```
\newpage
